---
title: "Creating themed escape rooms with escapeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating themed escape rooms with escapeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`escapeR` was first created inside an Ecological Statistics course. The bundled
rooms therefore use ecological examples: field counts, survey data, detection,
modelling, and reproducible reporting. The package architecture is more general
than that, however. An escape room is just a sequence of small tasks where
students use R to unlock the next step.

This vignette explains how to create rooms in any theme, such as medicine,
agriculture, social sciences, economics, psychology, or environmental policy. It
also shows how to bundle those rooms into a shareable room pack that can be
submitted to the `escapeR` GitHub repository.

## The room architecture

Every room is built from separate components:

- `id`: a short unique identifier, at most 8 characters, such as `"bpmean"` or
  `"yield"`.
- `module`: the broad teaching module, such as `"Medicine"` or
  `"Agriculture"`.
- `title`: the room title shown to students.
- `learning_goal`: what the student should practise.
- `introduction`: the story or setting.
- `challenge`: the task the student must solve.
- `hints`: one or more hints. Repeated calls to `hint()` reveal them in order.
- `correct_result`: the expected answer for simple numeric or text answers.
- `checker`: an optional custom function for more complex answer checking.
- `success`: the message shown after a correct answer.
- `failure`: an optional message shown after a wrong answer.

Most beginner rooms only need `correct_result`. For example, if the correct
answer is `130`, `escapeR` checks numeric submissions with a small tolerance. If
the correct answer is `"high"`, `escapeR` checks text after trimming spaces and
ignoring case. Use `checker` when you need more control.

## Step 1: choose a theme

Start with a familiar domain and a small learning goal. Good first rooms are
short, concrete, and assess one idea at a time.

Examples:

- Medicine: calculate an average blood pressure, count high-risk patients, or
  interpret a positive/negative test result.
- Agriculture: calculate mean crop yield, compare fertilizer treatments, or
  count plots affected by drought.
- Social sciences: calculate average survey satisfaction, tabulate responses,
  or classify whether an unemployment rate increased.

For a first contribution, aim for two or three rooms. That is enough to create a
small story arc while keeping the pull request easy to review.

## Step 2: create individual rooms

The example below creates a tiny medicine-themed escape room. The first room
asks students to create a vector and calculate a mean. The second asks them to
use a logical comparison and submit a category. The third asks for a simple
proportion.

```{r}
library(escapeR)

bpmean <- new_room(
  id = "bpmean",
  module = "Medicine",
  title = "The Clinic Intake",
  learning_goal = "Create a numeric vector and calculate its mean.",
  introduction = paste(
    "A clinic intake sheet lists systolic blood pressure readings for",
    "three patients seen before lunch."
  ),
  challenge = paste(
    "Create a vector with the values 120, 130, and 140.",
    "What is the average systolic blood pressure? Submit the mean."
  ),
  hints = c(
    "Use c() to combine the three readings into one vector.",
    "Use mean() to calculate the average of a numeric vector."
  ),
  correct_result = 130,
  success = "The intake sheet is complete, and the first cabinet opens.",
  failure = "Not quite. Check that all three readings are included before taking the mean."
)

riskcat <- new_room(
  id = "riskcat",
  module = "Medicine",
  title = "The Risk Label",
  learning_goal = "Use a logical comparison to classify a value.",
  introduction = paste(
    "A label printer waits beside the triage desk.",
    "It needs the correct risk category for the next patient."
  ),
  challenge = paste(
    "A systolic blood pressure of 145 is considered high if it is above 140.",
    "Submit high or normal."
  ),
  hints = c(
    "Ask R whether 145 > 140.",
    "If the comparison is TRUE, the category requested by the room is high."
  ),
  correct_result = "high",
  success = "The label prints clearly, and the triage desk unlocks."
)

posprop <- new_room(
  id = "posprop",
  module = "Medicine",
  title = "The Test Result Board",
  learning_goal = "Calculate a proportion from counts.",
  introduction = paste(
    "A result board shows 8 positive tests out of 40 tests performed.",
    "The ward door asks for the positive proportion."
  ),
  challenge = "Submit the proportion of tests that were positive.",
  hints = c(
    "A proportion is part divided by whole.",
    "In R, calculate 8 / 40."
  ),
  correct_result = 0.2,
  success = "The board accepts the proportion, and the ward door opens."
)
```

Notice that each room is self-contained. The student sees the introduction and
challenge, but the correct result and hints are stored separately. This makes
rooms easier to review, edit, translate, and reuse.

## Step 3: bundle rooms into a pack

A room pack is the unit that is easiest to share. It contains a pack ID, a short
description, the rooms, and one or more named escape sequences. A sequence is
just an ordered vector of room IDs.

```{r}
medicine_pack <- new_room_pack(
  id = "medpack",
  title = "Introductory medicine rooms",
  description = paste(
    "A small set of medicine-themed rooms for practising vectors,",
    "means, comparisons, and proportions."
  ),
  rooms = list(bpmean, riskcat, posprop),
  escapes = list(
    medmini = c("bpmean", "riskcat"),
    medfull = c("bpmean", "riskcat", "posprop")
  )
)
```

The pack can contain more than one escape sequence. In this example, `medmini`
is a two-room activity and `medfull` uses all three rooms.

## Step 4: test locally

Register the pack in your R session, inspect the rooms and escape sequences, and
then start a test game.

```{r}
register_room_pack(medicine_pack, replace = TRUE)

list_rooms()
list_escapes()

medicine_escape <- build_escape("medfull")
medicine_escape$room_ids
```

To play it interactively:

```{r, eval = FALSE}
escape(player = "demo_medicine", reset = TRUE, escape = medicine_escape)

# Room 1 answer
submit(130)

# Room 2 answer
submit("high")

# Room 3 answer
submit(0.2)
```

When developing a pack, test wrong answers as well. Good failure messages help
students recover without giving away the whole solution.

## Step 5: use a custom checker when needed

Sometimes a room has several acceptable answers. For example, a social-science
room might accept `"increase"`, `"increased"`, or `"up"`. In that case, provide a
`checker` function instead of `correct_result`.

```{r}
trend <- new_room(
  id = "trend",
  module = "Social sciences",
  title = "The Survey Trend",
  learning_goal = "Interpret a direction of change.",
  introduction = "A survey dashboard compares this year with last year.",
  challenge = "Satisfaction rose from 62 to 70 percent. Submit the direction of change.",
  hints = c(
    "Compare the second value with the first.",
    "Several words can describe an upward change."
  ),
  checker = function(answer) {
    tolower(trimws(as.character(answer))) %in% c("increase", "increased", "up", "rose")
  },
  success = "The dashboard accepts the trend.",
  failure = "Try describing whether the second value is higher or lower than the first."
)
```

Custom checkers should be short and predictable. They must return `TRUE` only
when the answer should unlock the room.

## Step 6: prepare a GitHub contribution

To submit rooms for integration into `escapeR`, place the room pack in a single
R file. A good file name uses the pack theme, for example:

```text
R/room_pack_medicine.R
R/room_pack_agriculture.R
R/room_pack_social.R
```

The file should define one function that returns the pack. For the medicine
example, the submitted file would look like this:

```{r, eval = FALSE}
medicine_room_pack <- function() {
  bpmean <- new_room(
    id = "bpmean",
    module = "Medicine",
    title = "The Clinic Intake",
    learning_goal = "Create a numeric vector and calculate its mean.",
    introduction = paste(
      "A clinic intake sheet lists systolic blood pressure readings for",
      "three patients seen before lunch."
    ),
    challenge = paste(
      "Create a vector with the values 120, 130, and 140.",
      "What is the average systolic blood pressure? Submit the mean."
    ),
    hints = c(
      "Use c() to combine the three readings into one vector.",
      "Use mean() to calculate the average of a numeric vector."
    ),
    correct_result = 130,
    success = "The intake sheet is complete, and the first cabinet opens.",
    failure = "Not quite. Check that all three readings are included before taking the mean."
  )

  riskcat <- new_room(
    id = "riskcat",
    module = "Medicine",
    title = "The Risk Label",
    learning_goal = "Use a logical comparison to classify a value.",
    introduction = "A label printer waits beside the triage desk.",
    challenge = paste(
      "A systolic blood pressure of 145 is considered high if it is above 140.",
      "Submit high or normal."
    ),
    hints = c(
      "Ask R whether 145 > 140.",
      "If the comparison is TRUE, the category requested by the room is high."
    ),
    correct_result = "high",
    success = "The label prints clearly, and the triage desk unlocks."
  )

  posprop <- new_room(
    id = "posprop",
    module = "Medicine",
    title = "The Test Result Board",
    learning_goal = "Calculate a proportion from counts.",
    introduction = paste(
      "A result board shows 8 positive tests out of 40 tests performed.",
      "The ward door asks for the positive proportion."
    ),
    challenge = "Submit the proportion of tests that were positive.",
    hints = c(
      "A proportion is part divided by whole.",
      "In R, calculate 8 / 40."
    ),
    correct_result = 0.2,
    success = "The board accepts the proportion, and the ward door opens."
  )

  new_room_pack(
    id = "medpack",
    title = "Introductory medicine rooms",
    description = paste(
      "A small set of medicine-themed rooms for practising vectors,",
      "means, comparisons, and proportions."
    ),
    rooms = list(bpmean, riskcat, posprop),
    escapes = list(
      medmini = c("bpmean", "riskcat"),
      medfull = c("bpmean", "riskcat", "posprop")
    )
  )
}
```

Then add a small test file, for example `tests/testthat/test-room-pack-medicine.R`:

```{r, eval = FALSE}
test_that("medicine room pack can be registered and played", {
  pack <- medicine_room_pack()
  register_room_pack(pack, replace = TRUE)

  expect_true(all(c("bpmean", "riskcat", "posprop") %in% list_rooms()$id))
  expect_equal(build_escape("medfull")$room_ids, c("bpmean", "riskcat", "posprop"))
  expect_true(pack$rooms[[1]]$checker(130))
  expect_true(pack$rooms[[2]]$checker("high"))
  expect_true(pack$rooms[[3]]$checker(0.2))
})
```

In a pull request, include:

1. The room-pack file.
2. A short test file.
3. Any small data files, if needed, under `inst/extdata/`.
4. A short README or vignette paragraph if the pack introduces a new teaching
   theme.

If the pull request is accepted, the pack can be integrated into the package by
keeping the submitted room-pack function and registering the pack from package
code or examples. The room IDs and escape sequence IDs then become available to
other users through `list_rooms()`, `list_escapes()`, and `build_escape()`.

## Design advice

Write rooms for learners, not for answer keys. A useful room should:

- practise one main R idea;
- use a theme-specific story but keep the statistical task clear;
- hide solution mechanics in hints rather than the challenge text;
- include two or three progressive hints;
- have a precise expected result or a very simple checker;
- give a supportive failure message when the likely mistake is predictable;
- use short stable IDs that will not clash with other packs.

The ecology rooms bundled with `escapeR` are one theme. The same architecture can
support medicine, agriculture, social sciences, and many other domains, as long
as each room remains a small, testable R learning step.
