---
title: "Creating themed escape rooms with escapeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating themed escape rooms with escapeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`escapeR` was first created for an Ecological Statistics course. The bundled
rooms therefore use ecological examples: field counts, survey data, detection,
modelling, and reproducible reporting. The package architecture is more general
than that, however. An escape room is just a sequence of small tasks where
students use R to unlock the next step.

This vignette explains how to create rooms in any theme, such as medicine,
agriculture, social sciences, economics, psychology, or environmental policy. It
also shows how to bundle those rooms into a shareable room pack that can be
submitted to the `escapeR` GitHub repository.

New rooms and quests are welcome. To contribute them to `escapeR`, submit a
GitHub pull request containing the room-pack definition, its registration entry,
tests, and any data files it needs. The complete process is described below.

## The room architecture

Every room is built from eleven components, although some are optional:

- `id`: a short unique identifier, at most 8 characters, such as `"bpmean"` or
  `"yield"`.
- `module`: the broad teaching module, such as `"Medicine"` or
  `"Agriculture"`.
- `title`: the room title shown to students.
- `learning_goal`: what the student should practise.
- `introduction`: the story or setting.
- `challenge`: the task the student must solve.
- `hints`: one or more hints. Repeated calls to `hint()` reveal them in order.
- `correct_result`: the expected answer for simple numeric or text answers.
- `checker`: an optional custom function for more complex answer checking.
- `success`: the message shown after a correct answer.
- `failure`: an optional message shown after a wrong answer.

Most beginner rooms only need a simple `correct_result`. For example, if the correct answer is `130`, `escapeR` checks numeric submissions with a small tolerance. If the correct answer is `"high"`, `escapeR` checks text after trimming spaces and ignoring case. Use `checker` instead of `correct_result` when you need more control.

## Step 1: choose a theme

Start with a familiar domain and a small learning goal. Good first rooms are
short, concrete, and assess one idea at a time.

Examples:

- Medicine: calculate an average blood pressure, count high-risk patients, or
  interpret a positive/negative test result.
- Agriculture: calculate mean crop yield, compare fertilizer treatments, or
  count plots affected by drought.
- Social sciences: calculate average survey satisfaction, tabulate responses,
  or classify whether an unemployment rate increased.

For a first contribution, aim for two or three rooms. That is enough to create a
small story arc while keeping the pull request easy to review.

## Step 2: create individual rooms

The example below creates a tiny medicine-themed escape room. The first room
asks students to create a vector and calculate a mean. The second asks them to
use a logical comparison and submit a category. The third asks for a simple
proportion.

```{r}
library(escapeR)

bpmean <- new_room(
  id = "bpmean",
  module = "Medicine",
  title = "The Clinic Intake",
  learning_goal = "Create a numeric vector and calculate its mean.",
  introduction = paste(
    "A clinic intake sheet lists systolic blood pressure readings for",
    "three patients seen before lunch."
  ),
  challenge = paste(
    "Create a vector with the values 120, 124, 111, 182, 130, and 145.",
    "What is the average systolic blood pressure? Submit the mean, rounded to 1 decimal."
  ),
  hints = c(
    "Use c() to combine multiple readings into one vector.",
    "Use mean() to calculate the average of a numeric vector."
  ),
  correct_result = 133.3,
  success = "The intake sheet is complete, and the first cabinet opens.",
  failure = "Not quite, young doctor. Check that all readings are included before taking the mean."
)

riskcat <- new_room(
  id = "riskcat",
  module = "Medicine",
  title = "The Risk Label",
  learning_goal = "Use a logical comparison to classify a value.",
  introduction = paste(
    "A label printer waits beside the triage desk.",
    "It needs the correct risk category for the next patient."
  ),
  challenge = paste(
    "A systolic blood pressure of 145 is considered high if it is above 140.",
    "Submit high or normal."
  ),
  hints = c(
    "Ask R whether 145 > 140.",
    "If the comparison is TRUE, the category requested by the room is high."
  ),
  correct_result = "high",
  success = "The label prints clearly, and the triage desk unlocks."
)

posprop <- new_room(
  id = "posprop",
  module = "Medicine",
  title = "The Test Result Board",
  learning_goal = "Calculate a proportion from counts.",
  introduction = paste(
    "A result board shows 8 positive tests out of 40 tests performed.",
    "The ward door asks for the positive proportion."
  ),
  challenge = "Submit the proportion of tests that were positive.",
  hints = c(
    "A proportion is part divided by whole.",
    "In R, calculate 8 / 40."
  ),
  correct_result = 0.2,
  success = "The board accepts the proportion, and the ward door opens."
)
```

Notice that each authoring object is self-contained. It contains everything
needed to review, edit, translate, and reuse the room. During play, `escapeR`
exposes the introduction, challenge, learning goal, and hints, but keeps the
correct result and checker out of the playable room object.

## Step 3: bundle rooms into a pack

A room pack is the unit that is easiest to share. It contains a pack ID, a short
description, the rooms, and one or more named escape sequences. A sequence is
just an ordered vector of room IDs.

```{r}
medicine_pack <- new_room_pack(
  id = "medpack",
  title = "Introductory medicine rooms",
  description = paste(
    "A small set of medicine-themed rooms for practising vectors,",
    "means, comparisons, and proportions."
  ),
  rooms = list(bpmean, riskcat, posprop),
  escapes = list(
    medmini = c("bpmean", "riskcat"),
    medfull = c("bpmean", "riskcat", "posprop")
  )
)
```

The pack can contain more than one escape sequence. In this example, `medmini`
is a two-room activity and `medfull` uses all three rooms.

## Step 4: test locally

Before submitting a pull request, register the pack in your R session, inspect
the rooms and escape sequences, and then start a test game.

```{r}
register_room_pack(medicine_pack, replace = TRUE)

list_rooms()
list_escapes()

medicine_escape <- build_escape("medfull")
medicine_escape$room_ids
```

To play it interactively:

```{r, eval = FALSE}
escape(player = "demo_medicine", reset = TRUE, escape = medicine_escape)

# Room 1 answer
submit(133.3)

# Room 2 answer
submit("high")

# Room 3 answer
submit(0.2)
```

When developing a pack, test wrong answers as well as correct ones. Where
relevant, try numeric values, text, and the kinds of R objects your checker is
designed to accept. Good failure messages help students recover and may offer a
small nudge without giving away the solution.

## Step 5: use a custom checker when needed

Sometimes a room has several acceptable answers. For example, a social-science
room might accept `"increase"`, `"increased"`, or `"up"`. In that case, provide a
`checker` function instead of `correct_result`.

```{r}
trend <- new_room(
  id = "trend",
  module = "Social sciences",
  title = "The Survey Trend",
  learning_goal = "Interpret a direction of change.",
  introduction = "A survey dashboard compares this year with last year.",
  challenge = "Satisfaction rose from 62 to 70 percent. Submit the direction of change.",
  hints = c(
    "Compare the second value with the first.",
    "Several words can describe an upward change."
  ),
  checker = function(answer) {
    tolower(trimws(as.character(answer))) %in% c("increase", "increased", "up", "rose")
  },
  success = "The dashboard accepts the trend.",
  failure = "Try describing whether the second value is higher or lower than the first."
)
```

Custom checkers should be short and predictable. They must return `TRUE` only
when the answer should unlock the room.

## Step 6: prepare a GitHub contribution

An accepted contribution must do more than define a pack: it must also add the
pack to the package catalogue and test the student-facing submission path. The
steps below describe the complete route from a local room pack to rooms and
named escapes that are available immediately after `library(escapeR)`.

### 6.1 Add the room-pack factory

Place the pack in a single R source file. A good file name uses the pack theme,
for example:

```text
R/room-pack-medicine.R
R/room-pack-agriculture.R
R/room-pack-social.R
```

The file should define one function that takes no arguments and returns the
pack. The function is an internal package factory, so it does not need to be
exported. For the medicine example, the submitted file would look like this:

```{r, eval = FALSE}
medicine_room_pack <- function() {
  bpmean <- new_room(
    id = "bpmean",
    module = "Medicine",
    title = "The Clinic Intake",
    learning_goal = "Create a numeric vector and calculate its mean.",
    introduction = paste(
      "A clinic intake sheet lists systolic blood pressure readings for",
    "three patients seen before lunch."
  ),
  challenge = paste(
      "Create a vector with the values 120, 124, 111, 182, 130, and 145.",
      "What is the average systolic blood pressure? Submit the mean, rounded to 1 decimal."
  ),
  hints = c(
      "Use c() to combine multiple readings into one vector.",
      "Use mean() to calculate the average of a numeric vector."
  ),
    correct_result = 133.3,
    success = "The intake sheet is complete, and the first cabinet opens.",
    failure = "Not quite, young doctor. Check that all readings are included before taking the mean."
  )

  riskcat <- new_room(
    id = "riskcat",
    module = "Medicine",
    title = "The Risk Label",
    learning_goal = "Use a logical comparison to classify a value.",
    introduction = "A label printer waits beside the triage desk.",
    challenge = paste(
      "A systolic blood pressure of 145 is considered high if it is above 140.",
      "Submit high or normal."
    ),
    hints = c(
      "Ask R whether 145 > 140.",
      "If the comparison is TRUE, the category requested by the room is high."
    ),
    correct_result = "high",
    success = "The label prints clearly, and the triage desk unlocks."
  )

  posprop <- new_room(
    id = "posprop",
    module = "Medicine",
    title = "The Test Result Board",
    learning_goal = "Calculate a proportion from counts.",
    introduction = paste(
      "A result board shows 8 positive tests out of 40 tests performed.",
      "The ward door asks for the positive proportion."
    ),
    challenge = "Submit the proportion of tests that were positive.",
    hints = c(
      "A proportion is part divided by whole.",
      "In R, calculate 8 / 40."
    ),
    correct_result = 0.2,
    success = "The board accepts the proportion, and the ward door opens."
  )

  new_room_pack(
    id = "medpack",
    title = "Introductory medicine rooms",
    description = paste(
      "A small set of medicine-themed rooms for practising vectors,",
      "means, comparisons, and proportions."
    ),
    rooms = list(bpmean, riskcat, posprop),
    escapes = list(
      medmini = c("bpmean", "riskcat"),
      medfull = c("bpmean", "riskcat", "posprop")
    )
  )
}
```

Room IDs must be unique across all bundled and contributed rooms. Pack IDs and
named escape IDs must likewise avoid existing registered IDs. Keep every ID to
at most eight letters, numbers, or underscores, beginning with a letter.

### 6.2 Add the pack to the package catalogue

Defining `medicine_room_pack()` is not enough to make its rooms available to
users. Add a call to the factory inside `.bundled_room_packs()` in
`R/bundled-packs.R`:

```r
.bundled_room_packs <- function() {
  list(
    medicine_room_pack()
  )
}
```

If the catalogue already contains packs, retain them and add the new factory as
another list element:

```r
.bundled_room_packs <- function() {
  list(
    existing_room_pack(),
    medicine_room_pack()
  )
}
```

When `escapeR` loads, its `.onLoad()` hook registers every pack in this list.
Consequently, users do not need to call `register_room_pack()` for a bundled
contribution: its rooms appear in `list_rooms()`, its named sequences appear in
`list_escapes()`, and those sequences work with `build_escape()` immediately.

Do not add a registration call to an example or vignette as a substitute for
the catalogue entry. Such a call affects only the R session in which that
example happens to run.

### 6.3 Add any supporting data

If a room needs a small data file, add it under `inst/extdata/` and locate it in
the challenge with `escapeR_file()`. Use a distinctive file name to avoid
collisions. The room and its tests must not depend on files that exist only on
the contributor's computer.

For example, a room could direct students to read:

```r
read.csv(escapeR_file("medicine-example.csv"))
```

Keep contributed data small, document their origin and licence in the pull
request, and include only data that the package is permitted to redistribute.

### 6.4 Add end-to-end tests

Add a focused test file such as
`tests/testthat/test-room-pack-medicine.R`. Tests should verify permanent
registration and should submit answers through the exported game interface.
That is more useful than calling a room's internal checker directly because it
also tests registration, escape construction, room order, progress, and
submission handling.

For example:

```{r, eval = FALSE}
test_that("medicine room pack is registered and playable", {
  expect_true(all(
    c("bpmean", "riskcat", "posprop") %in% list_rooms()$id
  ))
  expect_true(all(c("medmini", "medfull") %in% list_escapes()$id))

  medicine_escape <- build_escape("medfull")
  expect_equal(
    medicine_escape$room_ids,
    c("bpmean", "riskcat", "posprop")
  )

  player <- paste0("medicine_test_", Sys.getpid())
  escape(player = player, reset = TRUE, escape = medicine_escape)

  expect_false(submit(100))
  expect_true(submit(133.3))
  expect_true(submit(" HIGH "))
  expect_true(submit(0.2))
})
```

Use a unique test-player name so repeated test runs do not reuse another test's
progress. Test at least one incorrect answer and every correct answer in the
sequence. If a room uses a custom checker, include representative accepted and
rejected submissions for that checker.

The tests use `testthat`. Its expectations describe the required behaviour:

```{r, eval = FALSE}
expect_true(x > 0)             # a condition holds
expect_false(is.na(x))         # a condition does not hold
expect_equal(result, 4)        # two values are equal
expect_error(log("text"))      # an error is triggered
expect_warning(sqrt(-1))       # a warning is triggered
expect_type(x, "double")       # typeof(x) is "double"
```

Package tests belong under `tests/testthat/`. Run them locally with:

```{r, eval = FALSE}
devtools::test()
```

### 6.5 Render documentation and check the package

Run the full package checks before opening the pull request:

```{r, eval = FALSE}
devtools::document()
devtools::test()
devtools::check()
```

Also open the rendered vignettes and play the contributed escape once in a
fresh R session. Check that the story, hints, failure messages, room order, and
all required files make sense from a student's perspective.

### 6.6 Open the pull request

The pull request should contain:

1. The room-pack factory file.
2. The new entry in `R/bundled-packs.R`.
3. An end-to-end test file.
4. Any supporting data under `inst/extdata/`, with source and licensing details.
5. Documentation changes if the pack introduces a new workflow or teaching
   theme.

In the pull-request description, summarize the learning goals, list the new
room and escape IDs, explain any data provenance, and report the result of
`devtools::check()`. Reviewers can then evaluate both the teaching design and
the technical integration from one self-contained contribution.

## Design advice

Write rooms for learners, not for answer keys. A useful room should:

- practise one main R idea;
- use a theme-specific story but keep the statistical task clear;
- hide solution mechanics in hints rather than the challenge text;
- include two or three progressive hints;
- have a precise expected result or a very simple checker;
- give a supportive failure message when the likely mistake is predictable;
- use short stable IDs that will not clash with other packs.

The ecology rooms bundled with `escapeR` are one theme. The same architecture
can support other domains, as long as each room remains a small, testable R
learning step.
