---
title: "Getting started with escapeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with escapeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(escapeR)
```

`escapeR` is a small escape-room game for learning R through ecological
statistics. Each room gives you a thread of the whole story, a task, and a lock. You solve the task
with ordinary R commands, then submit the answer to move to the next room.

The bundled quest starts with R foundations and moves through data import,
visualisation, data manipulation, simple ecological modelling, distance-sampling
ideas, and reproducible workflows.

## Start your first quest

To start playing you simply need to load the package and call `escape()`:

```{r, eval = FALSE}
library(escapeR)
escape()
```

In an interactive R session, `escape()` asks for your player name. Use a short
name you can remember, because `escapeR` saves your progress under that name.

You can also provide your player name directly:

```{r, eval = FALSE}
escape(player = "ana")
```

The first room is then printed in the console. It includes:

- the room number and title;
- a short introduction;
- the task you need to solve;
- the learning goal for that room.

The task is solved outside the game prompt. Use R as you normally would: create
objects, inspect data, calculate values, make plots, or fit models. When you
think you have the answer, submit it.

## Submit an answer

Use `submit()` with the answer that should open the current lock:

```{r, eval = FALSE}
submit(70)
```

If the answer is correct, `escapeR` shows the success message and moves you to
the next room. If the answer is not correct, the room remains locked and you can
try again.

While numeric answers are submitted as is, text answers should be submitted as character strings:

```{r, eval = FALSE}
submit("negative")
submit(".qmd")
```

For simple text locks, `escapeR` ignores leading and trailing spaces and is not
case-sensitive. Numeric answers are checked with a small tolerance unless a room
uses a custom checker.

## Ask for a hint

If you get stuck, call `hint()`:

```{r, eval = FALSE}
hint()
```

Some rooms have more than one hint. Repeated calls reveal the hints in order:

```{r, eval = FALSE}
hint()
hint()
```

Hints are meant to nudge you toward the R idea rather than simply giving away
the answer. In a classroom, it is usually worth trying the task first, asking R
what objects you have created, and then requesting a hint if the lock is still
not opening.

## See the current room again

If the console has filled up with other work, call `play()`:

```{r, eval = FALSE}
play()
```

`play()` does not restart the game. It simply prints the current room again, so
you can reread the task and learning goal.

## Check your status

Use `status()` to see where you are:

```{r, eval = FALSE}
status()
```

This tells you the active player, how many rooms have been solved out of those in the game, and which room is current. It is useful during longer activities or when returning to the game after a break.

## Resume later

Progress is saved automatically for each player using
`tools::R_user_dir("escapeR", "data")`. To resume, load the package and call
`escape()` again with the same player name:

```{r, eval = FALSE}
library(escapeR)
escape(player = "ana")
```

If saved progress exists for that player, the game resumes from the current
room. If no saved progress exists, a new quest starts.

## Restart a quest

To restart the active player's quest from the beginning, call:

```{r, eval = FALSE}
reset_game()
```

You can also reset a named player:

```{r, eval = FALSE}
reset_game(player = "ana")
```

Or start again directly with `escape(reset = TRUE)`:

```{r, eval = FALSE}
escape(player = "ana", reset = TRUE)
```

Use resetting with care in class: it deliberately starts that player's progress
again from room 1, so all previous progress is lost.

## Find the bundled data files

Several rooms ask you to read or inspect files included with the package. Use
`escapeR_file()` to find them. As an example, if a data file was called "dataX.csv" you would use

```{r, eval=FALSE}
escapeR_file("dataX.csv")
```

For example, a room might ask you to read a CSV file and then inspect it. One option would then be

```{r, eval = FALSE}
d <- read.csv(escapeR_file("dataX.csv"))
head(d)
```

The package also provides a separate, small survey data set. `survey_counts()`
returns this data frame directly; it does not read or modify `dados1.csv`. Store
the returned data frame in an object before working with it:

```{r}
survey <- survey_counts()
names(survey)
head(survey)
sum(is.na(survey))
```

It has five named columns (`site`, `habitat`, `count`, `distance_m`, and
`detected`). Two values in `count` are `NA`, included deliberately for the data
quality exercise. By contrast, `dados1.csv` has four columns and no missing
values. The survey data set appears in several rooms about data quality,
summaries, modelling, and distance sampling.

## See available rooms

Use `list_rooms()` to inspect the bundled room sequence:

```{r}
list_rooms()
```

The `id` column is useful when building a shorter custom quest, where you can provide a list of rooms to be played as a separate quest. See next section and the corresponding dedicated vignette for details on how to create new rooms and quests.

## Play a shorter or custom quest

Instructors can build a quest from selected room IDs, as an example here, a mini 3-room quest:

```{r, eval = FALSE}
short_quest <- build_escape(c("console", "vector", "plotwin"))
escape(player = "demo_short", reset = TRUE, escape = short_quest)
```

The same `escape()` command starts the custom sequence. The only difference is
that the `escape` argument receives an escape sequence created with
`build_escape()`.

If room packs have been registered, `list_escapes()` shows named sequences:

```{r}
list_escapes()
```

Those named sequences can also be passed to `build_escape()`.

## Useful commands

Here is the core command set while playing:

```{r, eval = FALSE}
escape()       # start or resume a quest
play()         # show the current room again
hint()         # request the next hint
submit(70)     # submit an answer
status()       # check progress
reset_game()   # restart the active quest
list_rooms()   # inspect available rooms
```

The most important habit is to solve the room in ordinary R first. The game is
the lock; R is the key. Can you get out?
